Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-02-10 paper.qmd
Generate the true model for s-curve

shape_value_curve <- calculate_effective_shape_value(.data = tSNE_data,
                                                   x = tSNE1, y = tSNE2)

num_bins_vec <- 1:18 ## Number of bins along the x-axis

  model_object <- fit_high_d_model(training_data = training_data, nldr_df_with_id = tSNE_data, x = "tSNE1", y = "tSNE2", num_bins_x = num_bins_vec[i], shape_val = shape_value_curve,
                             is_bin_centroid = TRUE,
                             is_rm_lwd_hex = FALSE,
                             benchmark_to_rm_lwd_hex = NA,
                             is_avg_high_d = TRUE, column_start_text = "x")

  # pred_df_training <- predict_2d_embeddings(test_data = training_data, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = tSNE_data, x = "tSNE1", y = "tSNE2")

  eval_df_training <- generate_eval_df(data = data, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

MSE_df_1 <- eval_data_training |>
  dplyr::mutate(method = "tSNE")
```

@@

  # pred_df_training <- predict_2d_embeddings(test_data = training_data, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = UMAP_data, x = "UMAP1", y = "UMAP2")

  eval_df_training <- generate_eval_df(data = data, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

@@

MSE_df_2 <- eval_data_training |>
  dplyr::mutate(method = "UMAP")
```

shape_value_curve <- calculate_effective_shape_value(.data = PHATE_data,
                                                   x = PHATE1, y = PHATE2)

num_bins_vec <- 1:18 ## Number of bins along the x-axis

  model_object <- fit_high_d_model(training_data = training_data, nldr_df_with_id = PHATE_data, x = "PHATE1", y = "PHATE2", num_bins_x = num_bins_vec[i], shape_val = shape_value_curve,
                             is_bin_centroid = TRUE,
                             is_rm_lwd_hex = FALSE,
                             benchmark_to_rm_lwd_hex = NA,
                             is_avg_high_d = TRUE, column_start_text = "x")

  # pred_df_training <- predict_2d_embeddings(test_data = training_data, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = PHATE_data, x = "PHATE1", y = "PHATE2")

  eval_df_training <- generate_eval_df(data = data, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

MSE_df_3 <- eval_data_training |>
  dplyr::mutate(method = "PHATE")

@@

shape_value_curve <- calculate_effective_shape_value(.data = TriMAP_data,
                                                   x = TriMAP1, y = TriMAP2)

num_bins_vec <- 1:10 ## Number of bins along the x-axis

  model_object <- fit_high_d_model(training_data = training_data, nldr_df_with_id = TriMAP_data, x = "TriMAP1", y = "TriMAP2", num_bins_x = num_bins_vec[i], shape_val = shape_value_curve,
                             is_bin_centroid = TRUE,
                             is_rm_lwd_hex = FALSE,
                             benchmark_to_rm_lwd_hex = NA,
                             is_avg_high_d = TRUE, column_start_text = "x")

  # pred_df_training <- predict_2d_embeddings(test_data = training_data, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = TriMAP_data, x = "TriMAP1", y = "TriMAP2")

  eval_df_training <- generate_eval_df(data = data, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

MSE_df_4 <- eval_data_training |>
  dplyr::mutate(method = "TriMAP")

shape_value_curve <- calculate_effective_shape_value(.data = PaCMAP_data,
                                                   x = PaCMAP1, y = PaCMAP2)

num_bins_vec <- 1:18 ## Number of bins along the x-axis

  model_object <- fit_high_d_model(training_data = training_data, nldr_df_with_id = PaCMAP_data, x = "PaCMAP1", y = "PaCMAP2", num_bins_x = num_bins_vec[i], shape_val = shape_value_curve,
                             is_bin_centroid = TRUE,
                             is_rm_lwd_hex = FALSE,
                             benchmark_to_rm_lwd_hex = NA,
                             is_avg_high_d = TRUE, column_start_text = "x")

  # pred_df_training <- predict_2d_embeddings(test_data = training_data, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = PaCMAP_data, x = "PaCMAP1", y = "PaCMAP2")

  eval_df_training <- generate_eval_df(data = data, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

MSE_df_5 <- eval_data_training |>
  dplyr::mutate(method = "PaCMAP")

MSE_df <- dplyr::bind_rows(MSE_df_1, MSE_df_2, MSE_df_3, MSE_df_4, MSE_df_5)

MSE_df$method <- factor(MSE_df$method, levels = c("tSNE", "UMAP", "PHATE", "TriMAP", "PaCMAP"))

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 144, linetype="dashed", 
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 144, linetype="dashed", 
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

error_df_lg_1 <- read_rds("data/s_curve/s_curve_summary_lg_tsne.rds")
error_df_lg_2 <- read_rds("data/s_curve/s_curve_summary_lg_umap.rds")
error_df_lg_3 <- read_rds("data/s_curve/s_curve_summary_lg_phate.rds")
error_df_lg_4 <- read_rds("data/s_curve/s_curve_summary_lg_trimap.rds")
error_df_lg_5 <- read_rds("data/s_curve/s_curve_summary_lg_pacmap.rds")

error_df_lg <- dplyr::bind_rows(error_df_lg_1, error_df_lg_2, error_df_lg_3, error_df_lg_4, error_df_lg_5)

error_df_lg$method <- factor(error_df_lg$method, levels = c("tSNE", "UMAP", "PHATE", "TriMAP", "PaCMAP"))

## To draw with AIC
aic_plot_lg <- ggplot(error_df_lg |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_lg,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_vline(xintercept = 1.347, linetype="dashed", 
                color = "red", size=0.5) +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("benchmark value")
## Effective number of bins along x-axis

mse_plot_lg <- ggplot(error_df_lg |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_lg,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_vline(xintercept = 1.347, linetype="dashed", 
                color = "red", size=0.5) +
  geom_line() +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("benchmark value")

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the long edges. What is the effective benchmark value to remove the long edges? $1.347$ is the chosen benchmark value to remove long edges. The value is selected through a visual inspection of the triangular mesh is conducted to verify whether it successfully eliminates long edges."
#| label: fig-diagnosticpltScurvelgrm
##| out-width: 100%
#| fig-pos: H

aic_plot_lg + mse_plot_lg +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8)) 
```

@@

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0, linetype="dashed",
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("shift value")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 0, linetype="dashed",
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("shift value")

@@

The approach involves employing the K-nearest neighbors (KNN) algorithm to identify the nearest hexagonal bin centroid in the 2D space. Subsequently, the coordinates of this centroid are assigned as the low-dimensional predicted values for the test data in 2D space. It is noteworthy that traditional NLDR methods, such as tSNE, often lack a direct predict function, making our approach valuable for generating predicted values in the absence of such functionalities.

@@

Here, $n$ represents the number of bins, $p$ represents the dimensions, $x_{ij}$ is the actual high-dimensional data, and $C_{x_ij}$ is the predicted bin centroid data in high dimensions.

@@

### Model interpretation

@@

After considering the models constructed from the algorithm for each NLDR techniques and visualizing them in high-D space with the original data, shows that all the NLDR techniques capture the non-linearity of the S-curve (see videos of the models linked in @fig-modelScurve). But thorough examination of each model in high-D space shows some surprising findings. The model of PHATE is flattened and lack of capturing the width of S-curve (see videos of the model with PHATE linked in @fig-modelScurve). This nature of PHATE is also evicted by the considerably higher AIC and MSE values for PHATE as shown in @fig-diagnosticpltScurve. Furthermore, there are two edges that are not connected to any points in the model of TriMAP (see see videos of the model with TriMAP linked in @fig-modelScurve). This model shows how the patterns in 2D representation of TriMAP capture by the model (see @fig-nldervis (d)). As shown in @fig-nldervis (b), UMAP have four clusters in a curvilinear structure. However, the model reveals that even though UMAP shows a curvilinear in 2D, UMAP actually capture the structure of the S-curve without any disjoint in the model (see see videos of the model with UMAP linked in @fig-modelScurve). In the 2D layouts of tSNE and PaCMAP (see @fig-nldervis (a), (e)), there are some sparse areas that are captured by the models effectively (see see videos of the model with tSNE and PaCMAP linked in @fig-modelScurve).

@@

### Prediction {#sec-prediction}

In this context, "prediction" denotes the 2D embedding generated for the NLDR technique.
The methodology encompasses identifying the nearest averaged high-D points in the high-D space for the test data, by computing high-D Euclidean distances. As the averaged high-D point corresponds to the lifting of the 2D model, determining its nearest counterpart allows us to map its hexagonal bin centroid coordinates. Consequently, these centroid coordinates serve as the assigned low-dimensional predicted values for the test data in the 2D space.

@@

@@

# num_bins_umap_gau <- calculate_effective_x_bins(.data = UMAP_data_gau, x = UMAP1,
#                                                 cell_area = 1) #41
num_bins_umap_gau <- 26

shape_val_umap_gau <- calculate_effective_shape_value(.data = UMAP_data_gau,
                                                      x = UMAP1, y = UMAP2)
## To extract bin centroids
hexbin_data_object_umap_gau <- extract_hexbin_centroids(nldr_df = UMAP_data_gau, num_bins = num_bins_umap_gau, shape_val = shape_val_umap_gau, x = UMAP1, y = UMAP2)

@@

# num_bins_phate_gau <- calculate_effective_x_bins(.data = PHATE_data_gau, x = PHATE1,
#                                                  cell_area = 1)
num_bins_phate_gau <- 70

shape_val_phate_gau <- calculate_effective_shape_value(.data = PHATE_data_gau,
                                                       x = PHATE1, y = PHATE2)
## To extract bin centroids
hexbin_data_object_phate_gau <- extract_hexbin_centroids(nldr_df = PHATE_data_gau, num_bins = num_bins_phate_gau, shape_val = shape_val_phate_gau, x = PHATE1, y = PHATE2)

@@

num_bins_trimap_gau <- 21
shape_val_trimap_gau <- calculate_effective_shape_value(.data = TriMAP_data_gau,
                                                        x = TriMAP1, y = TriMAP2)
## To extract bin centroids
hexbin_data_object_trimap_gau <- extract_hexbin_centroids(nldr_df = TriMAP_data_gau, num_bins = num_bins_trimap_gau, shape_val = shape_val_trimap_gau, x = TriMAP1, y = TriMAP2)

@@

# num_bins_pacmap_gau <- calculate_effective_x_bins(.data = PaCMAP_data_gau, x = PaCMAP1,
#                                                   cell_area = 1) #30
num_bins_pacmap_gau <- 18

shape_val_pacmap_gau <- calculate_effective_shape_value(.data = PaCMAP_data_gau,
                                                        x = PaCMAP1, y = PaCMAP2)
## To extract bin centroids
hexbin_data_object_pacmap_gau <- extract_hexbin_centroids(nldr_df = PaCMAP_data_gau, num_bins = num_bins_pacmap_gau, shape_val = shape_val_pacmap_gau, x = PaCMAP1, y = PaCMAP2)

@@

  # pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = tSNE_data_gau, x = "tSNE1", y = "tSNE2")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

@@

  # pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "UMAP")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = UMAP_data_gau, x = "UMAP1", y = "UMAP2")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

@@

  # pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "PHATE")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = PHATE_data_gau, x = "PHATE1", y = "PHATE2")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

@@

  # pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "TriMAP")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = TriMAP_data_gau, x = "TriMAP1", y = "TriMAP2")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

@@

  # pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "PaCMAP")

  pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = PaCMAP_data_gau, x = "PaCMAP1", y = "PaCMAP2")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, num_bins = num_bins_vec[i], col_start = "x")

@@

-

@@

```{r}
#| warning: false
#| echo: false
num_bins_pbmc <- calculate_effective_x_bins(.data = UMAP_pbmc, x = UMAP1,
                                              cell_area = 1) ##23

num_bins_pbmc <- 23

shape_value_pbmc <- calculate_effective_shape_value(.data = UMAP_pbmc,
                                                      x = UMAP1, y = UMAP2) ## 0.8772751
shape_value_pbmc <- 0.8772751

>>>>

@@

  pred_df_training <- predict_2d_embeddings(test_data = training_data, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "UMAP")

  # pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = UMAP_data, x = "UMAP1", y = "UMAP2")

  eval_df_training <- generate_eval_df(data = data, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

@@

MSE_df_umap_s_curve <- eval_data_training |>
  mutate(data_type = "training",
         method = "UMAP")
```

@@

## To draw with AIC
aic_plot <- ggplot(MSE_df_umap_s_curve |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 144, linetype="dashed", 
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df_umap_s_curve |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 144, linetype="dashed", 
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

@@

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = MSE_df |> dplyr::filter(total_mse == min(total_mse)) |> dplyr::pull(shift), linetype="dashed",
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("shift value")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df |> dplyr::filter(method == "UMAP"), aes(x = shift,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = MSE_df |> dplyr::filter(total_mse == min(total_mse)) |> dplyr::pull(shift), linetype="dashed",
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("shift value")

@@

<!--In this context, "prediction" denotes the 2D embedding generated for the NLDR technique.
The methodology encompasses identifying the nearest averaged high-D points in the high-D space for the test data, by computing high-D Euclidean distances. As the averaged high-D point corresponds to the lifting of the 2D model, determining its nearest counterpart allows us to map its hexagonal bin centroid coordinates. Consequently, these centroid coordinates serve as the assigned low-dimensional predicted values for the test data in the 2D space.

The prediction approach involves employing the K-nearest neighbors (KNN) algorithm to identify the nearest hexagonal bin centroid in the 2D space. Subsequently, the coordinates of this centroid are assigned as the low-dimensional predicted values for the test data in 2D space.

The prediction approach involves employing the K-nearest neighbors (KNN) algorithm to identify the nearest high-D model point (averaged high-D point) for the new high-D data point. Also, the coordinates of the high-D model point serve as the predicted 2D embedings for the new high-D data point. 
It is noteworthy that traditional NLDR methods, such as tSNE, often lack a direct predict function, making our approach valuable for generating predicted values in the absence of such functionalities.
-->

The prediction approach involves employing the K-nearest neighbors (KNN) algorithm. In this method, which operates as an unsupervised classification problem, the nearest high-D model point (averaged high-D point) is identified for the new high-D data point. Once the nearest high-D model point is identified, its coordinates are used as the predicted 2D embeddings for the new high-D data point. This step is particularly valuable due to the limitations of certain NLDR techniques, like tSNE, which don't provide a straightforward method for predicting 2D embeddings. As a result, our approach presents a solution that can generate predicted 2D embeddings, irrespective of the NLDR technique employed, effectively addressing this gap in functionality.

@@

Here, $n$ represents the number of observations, $p$ represents the dimensions, $x_{ij}$ is the actual high-dimensional data, and $C_{x_ij}$ is the predicted bin centroid data in high dimensions.

@@

<!--
### Model interpretation

@@

After considering the models constructed from the algorithm for each NLDR techniques and visualizing them in high-D space with the original data, shows that all the NLDR techniques capture the non-linearity of the S-curve (see videos of the models linked in @fig-modelScurve). But thorough examination of each model in high-D space shows some surprising findings. The model of PHATE is flattened and lack of capturing the width of S-curve (see videos of the model with PHATE linked in @fig-modelScurve). This nature of PHATE is also evicted by the considerably higher AIC and MSE values for PHATE as shown in @fig-diagnosticpltScurve. Furthermore, there are two edges that are not connected to any points in the model of TriMAP (see see videos of the model with TriMAP linked in @fig-modelScurve). This model shows how the patterns in 2D representation of TriMAP capture by the model (see @fig-nldervis (d)). As shown in @fig-nldervis (b), UMAP have four clusters in a curvilinear structure. However, the model reveals that even though UMAP shows a curvilinear in 2D, UMAP actually capture the structure of the S-curve without any disjoint in the model (see see videos of the model with UMAP linked in @fig-modelScurve). In the 2D layouts of tSNE and PaCMAP (see @fig-nldervis (a), (e)), there are some sparse areas that are captured by the models effectively (see see videos of the model with tSNE and PaCMAP linked in @fig-modelScurve). -->

@@

@@

num_non_empty_bins_gau <- df_bin_centroids_tsne_gau$hexID |> length()

@@

# num_bins_umap_gau <- calculate_effective_x_bins(.data = UMAP_data_gau, x = UMAP1,
#                                                 cell_area = 1) #41

shape_val_umap_gau <- calculate_effective_shape_value(.data = UMAP_data_gau,
                                                      x = UMAP1, y = UMAP2)
num_bins_umap_gau <- find_non_empty_bins(nldr_df = UMAP_data_gau, x = "UMAP1", y = "UMAP2", shape_val = shape_val_umap_gau, non_empty_bins = num_non_empty_bins_gau)

## To extract bin centroids
hexbin_data_object_umap_gau <- extract_hexbin_centroids(nldr_df = UMAP_data_gau, num_bins = num_bins_umap_gau, shape_val = shape_val_umap_gau, x = UMAP1, y = UMAP2)

@@

# num_bins_phate_gau <- calculate_effective_x_bins(.data = PHATE_data_gau, x = PHATE1,
#                                                  cell_area = 1)

shape_val_phate_gau <- calculate_effective_shape_value(.data = PHATE_data_gau,
                                                       x = PHATE1, y = PHATE2)
num_bins_phate_gau <- find_non_empty_bins(nldr_df = PHATE_data_gau, x = "PHATE1", y = "PHATE2", shape_val = shape_val_phate_gau, non_empty_bins = num_non_empty_bins_gau)

## To extract bin centroids
hexbin_data_object_phate_gau <- extract_hexbin_centroids(nldr_df = PHATE_data_gau, num_bins = num_bins_phate_gau, shape_val = shape_val_phate_gau, x = PHATE1, y = PHATE2)

@@

shape_val_trimap_gau <- calculate_effective_shape_value(.data = TriMAP_data_gau,
                                                        x = TriMAP1, y = TriMAP2)

num_bins_trimap_gau <- find_non_empty_bins(nldr_df = TriMAP_data_gau, x = "TriMAP1", y = "TriMAP2", shape_val = shape_val_trimap_gau, non_empty_bins = num_non_empty_bins_gau)

## To extract bin centroids
hexbin_data_object_trimap_gau <- extract_hexbin_centroids(nldr_df = TriMAP_data_gau, num_bins = num_bins_trimap_gau, shape_val = shape_val_trimap_gau, x = TriMAP1, y = TriMAP2)

@@

# num_bins_pacmap_gau <- calculate_effective_x_bins(.data = PaCMAP_data_gau, x = PaCMAP1,
#                                                   cell_area = 1) #30

shape_val_pacmap_gau <- calculate_effective_shape_value(.data = PaCMAP_data_gau,
                                                        x = PaCMAP1, y = PaCMAP2)

num_bins_pacmap_gau <- find_non_empty_bins(nldr_df = PaCMAP_data_gau, x = "PaCMAP1", y = "PaCMAP2", shape_val = shape_val_pacmap_gau, non_empty_bins = num_non_empty_bins_gau)

## To extract bin centroids
hexbin_data_object_pacmap_gau <- extract_hexbin_centroids(nldr_df = PaCMAP_data_gau, num_bins = num_bins_pacmap_gau, shape_val = shape_val_pacmap_gau, x = PaCMAP1, y = PaCMAP2)

@@

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

@@

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "UMAP")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

@@

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "PHATE")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

@@

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "TriMAP")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

@@

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "PaCMAP")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

@@

@@

shape_value_pbmc <- calculate_effective_shape_value(.data = UMAP_pbmc,
                                                      x = UMAP1, y = UMAP2)

num_bins_vec <- 1:35 ## Number of bins along the x-axis

  model_object <- fit_high_d_model(training_data = training_data_pbmc, nldr_df_with_id = UMAP_pbmc, x = "UMAP1", y = "UMAP2", num_bins_x = num_bins_vec[i], shape_val = shape_value_pbmc,
                             is_bin_centroid = TRUE,
                             is_rm_lwd_hex = FALSE,
                             benchmark_to_rm_lwd_hex = NA,
                             is_avg_high_d = TRUE, column_start_text = "PC_")

  pred_df_training <- predict_2d_embeddings(test_data = training_data_pbmc, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "UMAP")

  eval_df_training <- generate_eval_df(data = training_data_pbmc, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "PC_")

MSE_df_pbmc <- eval_data_training |>
  dplyr::mutate(data_type = "training",
                method = "UMAP")
```
